---
title: "LUCIDus: Latent Unknown Clustering with Integrated Data"
author: "Cheng Peng"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LUCIDus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(LUCIDus)
```

## About LUCIDus

The **LUCIDus** package is aiming to provide researchers in the genetic epidemiology community with an integrative tool in R to obtain a joint estimation of latent or unknown clusters/subgroups with multi-omics data and phenotypic traits. 

This package is an implementation for the novel statistical method proposed in the research paper "A Latent Unknown Clustering Integrating Multi-Omics Data (LUCID) with Phenotypic Traits^[https://doi.org/10.1093/bioinformatics/btz667]" published by the *Bioinformatics*. LUCID improves the subtype classification which leads to better diagnostics as well as prognostics and could be the potential solution for efficient targeted treatments and successful personalized medicine.

## LUCIDus Functions and Model Fitting

Four main functions, including `est_lucid`, `boot_lucid`, `sem_lucid`, and `tune_lucid`, are currently available for model fitting and feature selection. The model outputs can be summarized and visualized using `summary_lucid` and `plot_lucid` respectively. Predictions could be made with `pred_lucid`.

Here are the descriptions of LUCIDus functions:

| Function | Description |
|:-----:|:---------------------|
| `est_lucid()` | Estimates latent clusters using multi-omics data with/without the outcome of interest, and producing an IntClust object; missing values in biomarker data (Z) are allowed |
| `boot_lucid()` | Latent cluster estimation through bootstrapping for statistical inference |
| `sem_lucid()` | Latent cluster estimation with supplemented E-M algorithm for statistical inference |
| `tune_lucid()` | Grid search for tuning parameters using parallel computing to determine an optimal choice of three tuning parameters with minimum model BIC |
| `summary_lucid()` | Summarizes the results of integrative clustering based on an `IntClust` object |
| `plot_lucid()` | Produces a Sankey diagram for the results of integrative clustering based on an `IntClust` object |
| `pred_lucid()` | Predicts latent clusters and outcomes with an `IntClust` object and new data |
| `def_initial()` | Defines initial values of model parameters in `est_lucid()`, `sem_lucid()` , and `tune_lucid()` fitting |
| `def_tune()` | Defines selection options and tuning parameters in `est_lucid()`, `sem_lucid()` , and `tune_lucid()` fitting |
| `def_tol()` | Defines tolerance settings in `est_lucid()`, `sem_lucid()` , and `tune_lucid()` fitting |

### Examples

For a simulated data with 10 genetic features (5 causal) and 4 biomarkers (2 causal)

- Integrative clustering without feature selection applying `est_lucid()`^[Parameter initials and stopping criteria can be specified using `def_initial()` and `def_tol()`]
```{r}
set.seed(10)
IntClusFit <- est_lucid(G=G1,Z=Z1,Y=Y1,K=2,family="binary",Pred=TRUE)
```

- Check important model outputs with `summary_lucid()`
```{r}
summary_lucid(IntClusFit)
```

- Visualize the results with a Sankey diagram via `plot_lucid()`
```{r fig.height=4, fig.width=7.25}
plot_lucid(IntClusFit)
```

- Run LUCID to select informative features
- Parallel grid-search with 27 combinations of tuning parameters using `tune_lucid()`
```{r}
GridSearch <- tune_lucid(G=G1, Z=Z1, Y=Y1, K=2, Family="binary", USEY = TRUE, NoCores = 2,
                         LRho_g = 0.008, URho_g = 0.012, NoRho_g = 3,
                         LRho_z_invcov = 0.04, URho_z_invcov = 0.06, NoRho_z_invcov = 3,
                         LRho_z_covmu = 90, URho_z_covmu = 100, NoRho_z_covmu = 3)
```

- Check the grid-search results
```{r}
GridSearch$Results
```

- Determine the best tuning parameters
```{r}
GridSearch$Optimal
```

- Run LUCID to select informative features^[Note tuning parameters in `est_lucid()` are defined by `def_tune()`]
```{r}
set.seed(0.01*0.05*95)
IntClusFit <- est_lucid(G=G1,Z=Z1,Y=Y1,K=2,family="binary",Pred=TRUE,
                        tunepar = def_tune(Select_G=TRUE,Select_Z=TRUE,
                                           Rho_G=0.01,Rho_Z_InvCov=0.05,Rho_Z_CovMu=95))
```

- Identify selected features
```{r}
summary_lucid(IntClusFit)$No0G; summary_lucid(IntClusFit)$No0Z
colnames(G1)[summary_lucid(IntClusFit)$select_G]; colnames(Z1)[summary_lucid(IntClusFit)$select_Z]
```

- Select the informative features
```{r}
if(!all(summary_lucid(IntClusFit)$select_G==FALSE)){
    G_select <- G1[,summary_lucid(IntClusFit)$select_G]
}
if(!all(summary_lucid(IntClusFit)$select_Z==FALSE)){
    Z_select <- Z1[,summary_lucid(IntClusFit)$select_Z]
}
```

- Re-fit with selected features
```{r}
set.seed(10)
IntClusFitFinal <- est_lucid(G=G_select,Z=Z_select,Y=Y1,K=2,family="binary",Pred=TRUE)
```

- Visualize the results
```{r fig.height=4, fig.width=7.25}
plot_lucid(IntClusFitFinal)
```

- Covariates can be easily adjusted in LUCID
```{r}
# Re-run the model with covariates in the G->X path
set.seed(10)
IntClusCoFit <- est_lucid(G=G1,CoG=CoG,Z=Z1,Y=Y1,K=2,family="binary",Pred=TRUE)
```

- Check clustering summary and visualize the results
```{r fig.height=4, fig.width=7.25}
# Check important model outputs
summary_lucid(IntClusCoFit)

# Visualize the results
plot_lucid(IntClusCoFit)
```

- Feature selection can be done with covariates which are forced to stay in the model
```{r fig.height=4, fig.width=7.25}
# Re-run feature selection with covariates in both paths
set.seed(10)
IntClusCoFit <- est_lucid(G=G1,CoG=CoG,Z=Z1,Y=Y1,CoY=CoY,K=2,family="binary",Pred=TRUE,
                          initial=def_initial(), itr_tol=def_tol(),
                          tunepar = def_tune(Select_G=TRUE,Select_Z=TRUE,
                          Rho_G=0.02,Rho_Z_InvCov=0.1,Rho_Z_CovMu=93))

# Re-fit with selected features with covariates
set.seed(10)
IntClusCoFitFinal <- est_lucid(G=G_select,CoG=CoG,Z=Z_select,Y=Y1,CoY=CoY,K=2,
                               family="binary",Pred=TRUE)

# Visualize the results
plot_lucid(IntClusCoFitFinal)
```

- Bootstrap method for variance estimation
```{r}
set.seed(10)
BootCIs <- boot_lucid(G = G1, CoG = CoG, Z = Z1, Y = Y1, CoY = CoY, 
                      useY = TRUE, family = "binary", K = 2, R=50)
BootCIs$Results
```

- Supplemented EM-algorithm for variance estimation
```{r eval = FALSE}
set.seed(10)
sem_lucid(G=G2,Z=Z2,Y=Y2,useY=TRUE,K=2,Pred=TRUE,family="normal",Get_SE=TRUE,
            itr_tol = def_tol(tol_b = 1e-02, tol_m = 1e-02, tol_s = 1e-02, tol_g = 1e-02))
```


## Final Remarks
- We recommend applying LUCIDus as a framework instead of separate functions;
- Extension to handle missingness in metabolomics data (Z) is on the way;
- An alternative option to obtain SE of parameter estimates is under testing.

## Acknowledgments
- David V. Conti, Ph.D.
- Zhao Yang, Ph.D.
- USC IMAGE Group^[Supported by the National Cancer Institute at the National Institutes of Health Grant P01 CA196569]

